\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2019-12-06 analysis/paper/paper.tex
Lit-Review Edits
TODO: fix table references in lit review

debug case study 3 after casebase pull request complete

To conduct our review on current methods available in survival analysis,
a scraper was coded to examine all packages in the survival task view on
CRAN (ref). There was a total of 259 packages, many of which are not
relevant to the problems casebase addresses. To filter these packages, a
scraper was used to search for competing risks, cumulative incidence
functions, non- proportional hazards and penalization. This reduced our
set of 259 packages down to 60. From here, we manually examined each
package to determine which ones are comparable to casebase, with either
a different approach to the same problems or future features we would
like to add to casebase.

@@

Many packages exist that implement semi-parametric cox models. Of those,
glmnet, glmpath, penalized, RiskRegression introduce penalization in
their model fitting. CoxRidge, rstpm2 and survival include penalization,
but they implement the extended cox model permitting intuitive non
proportional hazard modeling. In terms of penalization, the benefit of
casebase comes from its flexibility. Linear models have a wide tool set,
considerably larger than those made for cox regression (reference for
number of packages between their task views, regression is spread out
across Multivariate, SocialSciences, Robust among others. The focus of
this task view is on Optimization Infrastructure Packages , General
Purpose Continuous Solvers , Mathematical Programming Solvers , and
Specific Applications in Optimization , or Multi Objective Optimization
). Casebase permits many regularization tools developed for linear
regression to be repurposed within a survival context.

Parametric models permit non-proportional hazards and are implemented in
a handful of packages. Of interest are CFC, flexsurv, SmoothHazard,
Rsptm2 and Survival. Smooth hazard is limited to weibull distributions,
where as flexsurv and survival contain implementations for users to
supply any distribution they would like. Flexsurv, smooth hazard, mets
and rstpm2 can use splines, permitting a flexible fit over time. Though
casebase is limited to 3 distributions at the moment, it also uses
splines allowing for distributions that may not be consistent over time.

Of the methods mentioned so far, only CFC, survival, mets and flexsurv
contain implementations for competing risks(how is it done for
parametric models?). The differentiating factor between these packages
and casebase is that casebase handles competing risks through multiple
logistic regression.

A feature in Smooth hazard, survival and Rstpm2 are they they can handle
interval censoring, where as casebase currently can not.

Of all the packages listed in the table, CFC, Flexsurv, mets,
RiskRegression, rstpm2 and survival have implementations for cumulative
incidence. Since casebase is parametric, the baseline hazard is known
and requires integration to retrieve the cumulative incidence. This is
similar to the approaches of CFC, flexsurv,\\

@@

\begin{CodeInput}
R> #demonstrating incidence density before imputation
R> plot(popTime(as.data.frame(support),time = "d.time",event = "death"))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> #demonstrating incidence density after imputation
R> plot(popTime(as.data.frame(workingCompleteData),time = "d.time",event = "death"))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> #Using casebase's glmnet implementation, using splines and a case to base ratio of 100,
R> #to determine the hazard function
R> cb.Model=fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
R>                              formula_time = ~d.time,alpha=1,ratio=100)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> #Estimating the absolute risk curve using the newData parameter.
R> casebaseAbsolute=absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 0.5),newdata = t(newData) ,
R>                               s="lambda.1se",method=c("numerical"))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> #Create u and xCox, which will be used when fitting Cox with glmnet.
R> u=Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
R> xCox=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))[-c(sam),]
R> #hazard for cox using glmnet
R> coxglmFit=cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
R> #convergence demonstrated in plot
R> plot(coxglmFit)
R> #taking the coefficient estimates for later use
R> nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
R> nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
R> #creating a new dataset that only contains the covariates chosen through glmnet.
R> #cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
R> cleanCoxData<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,nonzero_covariate_cox$X1]))
R> #newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
R> #fitting a cox model using regular estimation, however we will not keep it.
R> #this is used more as an object place holder.
R> coxFit <- coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
R> #The coefficients of this object will be replaced with the estimates from coxglmFit.
R> #Doing so makes it so that everything is invalid aside from the coefficients.
R> #In this case, all we need to estimate the absolute risk is the coefficients.
R> #Std. error would be incorrect here, if we were to draw error bars.
R> coxFit$coefficients<-nonzero_coef_cox@x
R> #Fitting absolute risk curve for cox+glmnet
R> newDataCox=newData[nonzero_covariate_cox$X1-1]
R> 
R> abCoxFit<-survfit(coxFit,newdata=as.data.frame(t(newDataCox)),time = seq(0,max(completeData$d.time), 1),type="breslow")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> #creating a surv object to be used to fit an unadjusted absolute risk curve.
R> # (Kaplan-Meier risk curve)
R> km <- Surv(time = completeData$d.time, event = completeData$death)
R> abKm<-survfit(km~1,type='kaplan-meier',conf.type='log')
\end{CodeInput}
\end{CodeChunk}

@@

This report was generated on 2019-12-04 13:59:29 using the following
computational environment and dependencies:

@@

#>  date     2019-12-04                  

@@

#>  ! package       * version    date       lib
#>    abind           1.4-5      2016-07-21 [1]

@@

#>    bookdown        0.16       2019-11-22 [1]
#>    boot            1.3-23     2019-07-05 [1]

@@

#>    car             3.0-5      2019-11-15 [1]
#>    carData         3.0-3      2019-11-16 [1]

@@

#>  R casebasepaper * 0.0.0.9000 <NA>       [?]

@@

#>    class           7.3-15     2019-01-01 [1]

@@

#>    colorspace    * 1.4-1      2019-03-18 [1]

@@

#>    curl            4.2        2019-09-24 [1]
#>    data.table    * 1.12.6     2019-10-18 [1]

@@

#>    DEoptimR        1.0-8      2016-11-19 [1]

@@

#>    e1071           1.7-2      2019-06-05 [1]

@@

#>    foreign         0.8-72     2019-08-02 [1]

@@

#>    git2r           0.26.1     2019-06-29 [1]

@@

#>    here            0.1        2017-05-28 [1]

@@

#>    jomo            2.6-10     2019-10-22 [1]

@@

#>    laeken          0.5.0      2019-01-10 [1]
#>    lattice       * 0.20-38    2018-11-04 [1]
#>    lava            1.6.6      2019-08-01 [1]

@@

#>    lme4            1.1-21     2019-03-05 [1]
#>    lmtest          0.9-37     2019-04-30 [1]

@@

#>    MASS            7.3-51.4   2019-03-31 [1]

@@

#>    mice          * 3.6.0      2019-07-10 [1]
#>    minqa           1.2.4      2014-10-09 [1]
#>    mitml           0.3-7      2019-01-07 [1]

@@

#>    nloptr          1.2.1      2018-10-03 [1]
#>    nnet            7.3-12     2016-02-02 [1]
#>    numDeriv        2016.8-1.1 2019-06-06 [1]
#>    openxlsx        4.1.3      2019-11-07 [1]
#>    pan             1.6        2018-06-29 [1]

@@

#>    ranger          0.11.2     2019-03-07 [1]

@@

#>    rio             0.5.16     2018-11-26 [1]

@@

#>    robustbase      0.93-5     2019-05-12 [1]
#>    rpart           4.1-15     2019-04-12 [1]

@@

#>    sp              1.3-2      2019-11-07 [1]

@@

#>    timereg         1.9.4      2019-07-30 [1]

@@

#>    vcd             1.4-4      2017-12-06 [1]

@@

#>    VIM           * 4.8.0      2019-02-11 [1]

@@

#>    xtable          1.8-4      2019-04-21 [1]

@@

#>    zip             2.0.4      2019-09-01 [1]
#>    zoo             1.8-6      2019-05-28 [1]

@@

#>  <NA>                                    

@@

#>  R -- Package was removed from disk.

@@

\begin{CodeOutput}
#> Local:    master /Users/jislam/Desktop/Gits/cbpaper
#> Remote:   master @ origin (https://github.com/sahirbhatnagar/cbpaper.git)
#> Head:     [5022c53] 2019-12-04: rough litreview and casestudy 3 added
\end{CodeOutput}
\end{CodeChunk}

>>>>

To conduct our review on current methods available in survival analysis,
a scraper was coded to examine all packages in the survival task view on
CRAN \citeyearpar{survTaskView}. There was a total of 259 packages, many
of which are not relevant to the problems casebase addresses. To filter
these packages, a scraper was used to search for competing risks,
cumulative incidence functions, non- proportional hazards and
penalization. This reduced our set of 259 packages down to 60. From
here, we manually examined each package to determine which ones are
comparable to casebase, with either a different approach to the same
problems or future features we would like to add to casebase.

@@

Many packages exist that implement semi-parametric cox models. Of those,
glmnet \citeyearpar{regpathcox}???, glmpath \citeyearpar{park_hastie},
penalized \citeyearpar{l1penal}, RiskRegression
\citeyearpar{gerds_blanche} introduce penalization in their model
fitting. CoxRidge \citeyearpar{perperoglou} , rstpm2
\citeyearpar{clements_liu} and survival \citeyearpar{survival-package}
include penalization, but they implement the extended cox model
permitting intuitive non proportional hazard modeling while the previous
four do not. In terms of penalization, the benefit of casebase comes
from its flexibility. Linear models have a wide tool set, considerably
larger than those made for cox regression (\emph{reference for number of
packages between their task views, regression is spread out across
Multivariate, SocialSciences, Robust among others. The focus of this
task view is on Optimization Infrastructure Packages , General Purpose
Continuous Solvers , Mathematical Programming Solvers , and Specific
Applications in Optimization , or Multi Objective Optimization}).
Casebase permits many regularization tools developed for linear
regression to be repurposed within a survival context.

Parametric models permit non-proportional hazards and are implemented in
a handful of packages. Of interest are CFC
\citeyearpar{sharabiani_mahani_2019}, flexsurv \citeyearpar{flexsurv} ,
SmoothHazard \citeyearpar{smoothHazard}, Rsptm2
\citeyearpar{clements_liu} and Survival \citeyearpar{survival-package}.
Smooth hazard is limited to weibull
distributions\citeyearpar{smoothHazard}, where as flexsurv and survival
contain implementations for users to supply any distribution they would
like \citeyearpar{survival-package}, \citeyearpar{flexsurv}. Flexsurv,
smooth hazard, mets and rstpm2 can use splines, permitting a flexible
fit over time \citeyearpar{flexsurv}, \citeyearpar{smoothHazard},
\citeyearpar{scheike2014estimating}, \citeyearpar{clements_liu} . Though
casebase is limited to 3 distributions at the moment, it also uses
splines allowing for distributions that may not be consistent over
survival time.

Of the methods mentioned so far, only CFC,flexsurv, mets and survival
contain implementations for competing risks
\citeyearpar{sharabiani_mahani_2019},
\citeyearpar{flexsurv},\citeyearpar{scheike2014estimating}
,\citeyearpar{survival-package} (\emph{how is it done for parametric
models?}). The differentiating factor between these packages and
casebase is that casebase handles competing risks through multiple
logistic regression. Smooth hazard, survival and Rstpm2 can handle
interval censoring, where as casebase currently can not. Interval
censoring is simple feature to implement and is a potential improvement
to add to casebase. (There's also multistate models, which I haven't
mentioned in this paper)

Of all the packages listed in the table, CFC , Flexsurv, mets,
RiskRegression, rstpm2 and survival have implementations for cumulative
incidence \citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
\citeyearpar{scheike2014estimating}, \citeyearpar{gerds_blanche},
\citeyearpar{clements_liu}, \citeyearpar{survival-package}. Since
casebase is parametric, the baseline hazard is known and requires
integration to retrieve the cumulative incidence. This is similar to the
approaches of CFC and flexsurv \citeyearpar{sharabiani_mahani_2019},
\citeyearpar{flexsurv}, while the other packages use semi-parametric
approaches to retrieve the baseline after fitting
\citeyearpar{scheike2014estimating}, \citeyearpar{gerds_blanche}. rstpm2
and survival use both for their respective parametric and
semi-parametric model implementations \citeyearpar{clements_liu},
\citeyearpar{survival-package}.

@@

\begin{CodeInput}
R> #demonstrating incidence density before imputation
R> plot(casebase::popTime(as.data.frame(support),time = "d.time",event = "death"))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> #demonstrating incidence density after imputation
R> plot(casebase::popTime(as.data.frame(workingCompleteData),time = "d.time",event = "death"))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> #Using casebase's glmnet implementation, using splines and a case to base ratio of 100,
R> #to determine the hazard function
R> cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
R>                              formula_time = ~d.time,alpha=1,ratio=100)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> #Estimating the absolute risk curve using the newData parameter.
R> casebaseAbsolute=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 0.5),newdata = t(newData) ,
R>                               s="lambda.1se",method=c("numerical"))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> #Create u and xCox, which will be used when fitting Cox with glmnet.
R> u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
R> xCox=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))[-c(sam),]
R> #hazard for cox using glmnet
R> coxglmFit=glmnet::cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
R> #convergence demonstrated in plot
R> plot(coxglmFit)
R> #taking the coefficient estimates for later use
R> nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
R> nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
R> #creating a new dataset that only contains the covariates chosen through glmnet.
R> #cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
R> cleanCoxData<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,nonzero_covariate_cox$X1]))
R> #newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
R> #fitting a cox model using regular estimation, however we will not keep it.
R> #this is used more as an object place holder.
R> coxFit <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
R> #The coefficients of this object will be replaced with the estimates from coxglmFit.
R> #Doing so makes it so that everything is invalid aside from the coefficients.
R> #In this case, all we need to estimate the absolute risk is the coefficients.
R> #Std. error would be incorrect here, if we were to draw error bars.
R> coxFit$coefficients<-nonzero_coef_cox@x
R> #Fitting absolute risk curve for cox+glmnet
R> newDataCox=newData[nonzero_covariate_cox$X1-1]
R> 
R> abCoxFit<-survival::survfit(coxFit,newdata=as.data.frame(t(newDataCox)),time = seq(0,max(completeData$d.time), 1),type="breslow")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> #creating a surv object to be used to fit an unadjusted absolute risk curve.
R> # (Kaplan-Meier risk curve)
R> km <- survival::Surv(time = completeData$d.time, event = completeData$death)
R> abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
\end{CodeInput}
\end{CodeChunk}

@@

This report was generated on 2019-12-06 16:16:07 using the following
computational environment and dependencies:

@@

#>  date     2019-12-06                  

@@

#>  package     * version    date       lib

@@

@@

@@

@@

@@

#>  colorspace    1.4-1      2019-03-18 [1]

@@

#>  data.table    1.12.6     2019-10-18 [1]

@@

@@

@@

@@

@@

@@

@@

#>  lattice       0.20-38    2018-11-04 [1]

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

\begin{CodeOutput}
#> Local:    master /Users/jislam/Desktop/Gits/cbpaper
#> Remote:   master @ origin (https://github.com/Jesse-Islam/cbpaper.git)
#> Head:     [22bfed0] 2019-12-06: currentDraftEdits
\end{CodeOutput}
\end{CodeChunk}
